### 2017-10-03 chapters/introduction/mechanics.tex
Update figures. Change size and convert to eps

\subsection{Kinematics}
We represent the heart as a continuum body $\mathfrak{B}$ embedded in
$\mathbb{R}^3$. A configuration of $\mathfrak{B}$ is a mapping $\chi:
\mathfrak{B} \rightarrow \mathbb{R}^3$. 
We denote the \emph{reference configuration} of the heart by $\Omega
\equiv \chi_0(\mathfrak{B})$, and the \emph{current configuration} by $\omega
\equiv \chi(\mathfrak{B})$. The mapping $\varphi :  \Omega
\rightarrow \omega$, given by the composition $\varphi = \chi
\circ \chi_0^{-1}$, is a smooth, orientation preserving (positive
determinant) and invertible map. We denote the coordinates in the
reference configuration by $\Xvec \in \Omega$, and the coordinates in the current
configuration by $\xvec \in \omega$. The coordinates $\Xvec$ and $\xvec$ are
commonly referred to as material and spatial points respectively, and
are related through the mapping $\varphi$, by $\xvec = \varphi(\Xvec)$.
For time-dependent problems it is common to make  the time-dependence
explicitly by writing $\xvec = \varphi(\Xvec, t)$. In the following
we will only focus on the mapping between two configurations and
therefore no time-dependence is needed. The mapping $\varphi$ will be
referred to as the \emph{motion}. The \emph{deformation gradient} is a
rank-2 tensor, defined as the partial derivative of the motion with
respect to the material coordinates:
\begin{align}
  \F = \nabla_{\Xvec} \varphi= \Grad \xvec.
  \label{eq:deformation_gradient}
\end{align}
Here we also introduce the notation $\Grad$, which means derivative
with respect to reference coordinates.
The deformation gradient maps vectors in the reference configuration to
vectors in the current configuration, and belongs to the space of
linear transformations from $\mathbb{R}^3$ to $\mathbb{R}^3$ with
strictly positive determinant, which we denote by
$\mathrm{Lin}^+$. Another important quantity is the
\emph{displacement} field  
\begin{align}
  \uvec = \xvec-\Xvec, 
  \label{eq:displacement}
\end{align}
which relates positions in the reference configuration to positions
in the current configuration. From \eqref{eq:deformation_gradient} we
see that
\begin{align}
  \F = \Grad \xvec = \Grad \uvec + \Grad \Xvec = \Grad \uvec + \I.
\end{align}
Figure \ref{fig:motion} shows how a point in the reference
configuration is transformed to a point in the current configuration. 
\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.6\textwidth]{chapters/introduction/figures/motion.pdf}
\caption{Relations between the current and reference configuration.}
\label{fig:motion}
\end{figure}
Some other useful quantities are the \emph{right Cauchy-Green} deformation
tensor $\C = \F^T\F$, the \emph{left Cauchy-Green} deformation tensor
$\mathbf{B} = \F\F^T$, the \emph{Green-Lagrange} strain tensor
$\mathbf{E} = \frac{1}{2}(\C - \I)$, and the determinant of the
deformation gradient $J = \det \F$.

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{chapters/introduction/figures/boundaries}
\caption{Illustration of the different boundaries in a bi-ventricular
  domain.}
\label{fig:boundaries}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{chapters/introduction/figures/Hill_muscle_model.pdf}
\caption{The classical three-element Hill muscle model with one
  contractile element and two non-linear springs, one arranged in
  series and one parallel. }
\label{fig:hill_muscle_model}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.7\textwidth]{chapters/introduction/figures/actstrain.pdf}
\caption{Illustration of the active strain formulation. During the active
deformation, the sarcomeres shortens as if they were all detached. The
elastic deformation ensures compatibility of the tissue.}
\label{fig:actstrain}
\end{figure}

>>>>

\subsection{Kinematics}
We represent the heart as a continuum body $\mathfrak{B}$ embedded in
$\mathbb{R}^3$. A configuration of $\mathfrak{B}$ is a mapping $\chi:
\mathfrak{B} \rightarrow \mathbb{R}^3$. 
We denote the \emph{reference configuration} of the heart by $\Omega
\equiv \chi_0(\mathfrak{B})$, and the \emph{current configuration} by $\omega
\equiv \chi(\mathfrak{B})$. The mapping $\varphi :  \Omega
\rightarrow \omega$, given by the composition $\varphi = \chi
\circ \chi_0^{-1}$, is a smooth, orientation preserving (positive
determinant) and invertible map. We denote the coordinates in the
reference configuration by $\Xvec \in \Omega$, and the coordinates in the current
configuration by $\xvec \in \omega$. The coordinates $\Xvec$ and $\xvec$ are
commonly referred to as material and spatial points respectively, and
are related through the mapping $\varphi$, by $\xvec = \varphi(\Xvec)$.
For time-dependent problems it is common to make  the time-dependence
explicitly by writing $\xvec = \varphi(\Xvec, t)$. In the following
we will only focus on the mapping between two configurations and
therefore no time-dependence is needed. The mapping $\varphi$ will be
referred to as the \emph{motion}. The \emph{deformation gradient} is a
rank-2 tensor, defined as the partial derivative of the motion with
respect to the material coordinates:
\begin{align}
  \F = \nabla_{\Xvec} \varphi= \Grad \xvec.
  \label{eq:deformation_gradient}
\end{align}
Here we also introduce the notation $\Grad$, which means derivative
with respect to reference coordinates.
The deformation gradient maps vectors in the reference configuration to
vectors in the current configuration, and belongs to the space of
linear transformations from $\mathbb{R}^3$ to $\mathbb{R}^3$ with
strictly positive determinant, which we denote by
$\mathrm{Lin}^+$. Another important quantity is the
\emph{displacement} field  
\begin{align}
  \uvec = \xvec-\Xvec, 
  \label{eq:displacement}
\end{align}
which relates positions in the reference configuration to positions
in the current configuration. From \eqref{eq:deformation_gradient} we
see that
\begin{align}
  \F = \Grad \xvec = \Grad \uvec + \Grad \Xvec = \Grad \uvec + \I.
\end{align}

Some other useful quantities are the \emph{right Cauchy-Green} deformation
tensor $\C = \F^T\F$, the \emph{left Cauchy-Green} deformation tensor
$\mathbf{B} = \F\F^T$, the \emph{Green-Lagrange} strain tensor
$\mathbf{E} = \frac{1}{2}(\C - \I)$, and the determinant of the
deformation gradient $J = \det \F$.

@@

\begin{figure}[htbp]
  \centering
    \includegraphics{chapters/introduction/figures/boundaries}
\caption{Illustration of the different boundaries in a bi-ventricular
  domain.}
\label{fig:boundaries}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
    \includegraphics{chapters/introduction/figures/Hill_muscle_model}
\caption{The classical three-element Hill muscle model with one
  contractile element and two non-linear springs, one arranged in
  series and one parallel. }
\label{fig:hill_muscle_model}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
    \includegraphics{chapters/introduction/figures/actstrain}
\caption{Illustration of the active strain formulation. During the active
deformation, the sarcomeres shortens as if they were all detached. The
elastic deformation ensures compatibility of the tissue.}
\label{fig:actstrain}
\end{figure}

### 2017-10-03 chapters/introduction/physiology.tex
Update figures. Change size and convert to eps

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.7\textwidth]{chapters/introduction/figures/heart_anatomy.pdf}
\caption{Schematic representation of the heart's anatomy. (Adapted and modified from Wapcaplet in Sodipodi).}
\label{fig:heart_anatomy}
\end{figure}

@@

Each myocyte is composed of bundles of myofibriles, which again contains
\emph{myofilaments}. The myofilaments consist of a repeated pattern
of lines and bands which can be seen as a collection of individual
contractile units called \emph{sarcomeres}. A simple sketch of a
sarcomere is shown in Figure \ref{fig:sarcomere}. 
\begin{figure}[htbp]
  \centering
    \includegraphics[width=1.0\textwidth]{chapters/introduction/figures/Sarcomere.png}
\caption{Illustration of the sarcomere structure.}
\label{fig:sarcomere}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.85\textwidth]{chapters/introduction/figures/cardiac_cycle.png}
\caption{Showing the relation between pressure and volume in the left
  ventricle during the cardiac cycle. To the left we show a reduced version of
the classical Wiggers diagram with pressure and volume plotted against
time. To the right we show the pressure volume loop
with volume plotted on the $x-$axis and pressure on the $y-$axis. }
\label{fig:pv_loop}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{chapters/introduction/figures/pvr.pdf}
\caption{Illustration of the end diastolic pressure volume
  relationship (EDPVR) and the end systolic pressure volume
  relationship (ESPVR). The volume axis intercept $V_0$ from the time
  varying elastance model is also shown. }
\label{fig:intro_pvr}
\end{figure}

>>>>

\begin{figure}[htbp]
  \centering
    \includegraphics{chapters/introduction/figures/heart_anatomy}
\caption{Schematic representation of the heart's anatomy. (Adapted and modified from Wapcaplet in Sodipodi).}
\label{fig:heart_anatomy}
\end{figure}

@@

Each myocyte is composed of bundles of myofibriles, which again contains
\emph{myofilaments}. The myofilaments consist of a repeated pattern
of lines and bands which can be seen as a collection of individual
contractile units called \emph{sarcomeres}. A simple sketch of a
sarcomere is shown in Figure \ref{fig:sarcomere}. 
\begin{figure}[htbp]
  \centering
    \includegraphics{chapters/introduction/figures/Sarcomere}
\caption{Illustration of the sarcomere structure.}
\label{fig:sarcomere}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
    \includegraphics{chapters/introduction/figures/cardiac_cycle}
\caption{Showing the relation between pressure and volume in the left
  ventricle during the cardiac cycle. To the left we show a reduced version of
the classical Wiggers diagram with pressure and volume plotted against
time. To the right we show the pressure volume loop
with volume plotted on the $x-$axis and pressure on the $y-$axis. }
\label{fig:pv_loop}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
    \includegraphics{chapters/introduction/figures/pvr}
\caption{Illustration of the end diastolic pressure volume
  relationship (EDPVR) and the end systolic pressure volume
  relationship (ESPVR). The volume axis intercept $V_0$ from the time
  varying elastance model is also shown. }
\label{fig:intro_pvr}
\end{figure}

### 2017-10-03 pre_thesis/abstract.tex
Update figures. Change size and convert to eps

\begin{abstract}
  Here is the abstract
\end{abstract}

>>>>

\begin{abstract}
  In this thesis we will study variational data assimilation
  techniques applied to the field of patient-specific cardiac
  mechanics modeling. We develop tenchniques for creation of
  personalized mechanical similations of human hearts, and use these
  simulations to extract biomarkers related to cardiac contractility, stress
  and work. The tools developed in this thesis could potentially be used to assess
  additional mechanical information \emph{in vivo}, beyond the
  capabilities of modern imaging techniques.   
\end{abstract}

### 2017-10-03 pre_thesis/acknowledgements.tex
Update figures. Change size and convert to eps

\newenvironment{acknowledgements}%
    {\cleardoublepage\thispagestyle{empty}\null\vfill\begin{center}%
    \bfseries Acknowledgements\end{center}}%
    {\vfill\null}
        \begin{acknowledgements}
        These are the acknowledgements.
      \end{acknowledgements}

>>>>

\section*{Acknowledgements}

This is my thesis submitted for the degree Doctor Scientiarum,
submitted to the Department of Informatics, University of Oslo. The
work conducted in this thesis took place in the period from October
2014 to September 2017.\\
\indent First of all, I wish to deeply acknowledge the Center for Cardiological
Innovation (CCI) for providing the financial support for my PhD.\\
\indent I would especially like thank my main supervisors Joakim and Sam,
which both have been excellent in providing useful feedback, and
directions for my thesis. Joakim, thank you for allways taking time to
answer any question I may have, whether it is related to work or
private. Sam, thank you for setting up all the collaborations and for
guiding my work in the right direction.\\
\indent I would also like to thank Simula Research Laboratory for providing
an outstanding working environment. In my opinion, the including
management and the nice people at Simula, makes Simula the perfect
place to pursue a PhD. A special thank to the people from the Cardiac
Modeling group, in particular to Siri which I shared office with for most
of my PhD, and to Gabriel who I worked very close with during my first
year. \\
\indent I am also grateful to the CCI for giving me the opportunity to work
with, and learn from such exceptional medical researchers and for
providing an environment for interdisciplinary collaborations. In
particular I would like to thank, Hans Henrik, John and Espen for many
interesting and fruitful discussions. \\
\indent Finally, I would like to thank my family for always reminding me what
truly matters in life. Elisabeth, I am
deeply grateful for all the support you have given me during these years. 
% \end{abstract}

\vspace{5cm}
\noindent
Henrik Nicolay Topnes Finsberg\\
Oslo, September, 2017

### 2017-10-03 pre_thesis/list_of_papers.tex
Fix bug
Update figures. Change size and convert to eps

\textbf{Paper 1}
\hfill  \\ \noindent
Balaban, Gabriel, Henrik Finsberg, Hans Henrik Odland, Marie
E. Rognes, Stian Ross, Joakim Sundnes, and Samuel
Wall. \emph{High-resolution data assimilation of cardiac mechanics
  applied to a dyssynchronous ventricle.}  International journal for
numerical methods in biomedical engineering
(2017). doi:10.1002/cnm.2863
Status: published
\\\\
\noindent
\textbf{Paper 2} 
\hfill  \\ 
Henrik, Finsberg, Gabriel Balaban,  Stian Ross, Trine F. H\r{a}land,
Hans Henrik Odland, Joakim Sundnes, and Samuel
Wall. \emph{Estimating cardiac contraction through high resolution
  data assimilation of a personalized mechanical model.} Journal of
Computational Science (2017). doi:10.1016/j.jocs.2017.07.013
\\Status: published
\\\\
\noindent
\textbf{Paper 3}
\hfill  \\
Henrik Finsberg, Ce Xi,  Ju Le Tan,  Liang Zhong, Martin Genet, Joakim
Sundnes, Lik Chuan Lee, and Samuel T. Wall.
\emph{Personalized bi-ventricular mechanical analysis using
  gradient-based optimization.} International journal for
numerical methods in biomedical engineering
\\Status: submitted
\\\\
\noident
\textbf{Paper 4}
\hfill  \\
Henrik Finsberg, John Aalen,Camilla K.
Larsen, Joakim Sundnes, Otto A. Smiseth, and 
Samuel T. Wall.
\emph{Assessment of regional myocardial work via adjoint-based
  data assimilation.} Biomechanics and Modeling in Mechanobiology
\\Status: submitted

>>>>

\textbf{Paper 1}
\hfill  \\ \noindent
Gabriel Balban, Henrik Finsberg, Hans Henrik Odland, Marie
E. Rognes, Stian Ross, Joakim Sundnes, and Samuel
Wall. \emph{High-resolution data assimilation of cardiac mechanics
  applied to a dyssynchronous ventricle.}  International journal for
numerical methods in biomedical engineering
(2017). doi:10.1002/cnm.2863
Status: published
\\\\
\noindent
\textbf{Paper 2} 
\hfill  \\ 
Henrik Finsberg, Gabriel Balaban,  Stian Ross, Trine F. H\r{a}land,
Hans Henrik Odland, Joakim Sundnes, and Samuel
Wall. \emph{Estimating cardiac contraction through high resolution
  data assimilation of a personalized mechanical model.} Journal of
Computational Science (2017). doi:10.1016/j.jocs.2017.07.013
\\Status: published
\\\\
\noindent
\textbf{Paper 3}
\hfill  \\
Henrik Finsberg, Ce Xi,  Ju Le Tan,  Liang Zhong, Martin Genet, Joakim
Sundnes, Lik Chuan Lee, and Samuel T. Wall.
\emph{Personalized bi-ventricular mechanical analysis using
  gradient-based optimization.} International journal for
numerical methods in biomedical engineering
\\Status: submitted
\\\\
\noindent
\textbf{Paper 4}
\hfill  \\
Henrik Finsberg, John Aalen,Camilla K.
Larsen, Joakim Sundnes, Otto A. Smiseth, and 
Samuel T. Wall.
\emph{Assessment of regional myocardial work from a
patient-specific cardiac mechanics model.} 
\\Status: to be submitted
